---
title: "Lecture_5Rcode"
author: "Nemanja Vaci"
date: "March 10, 2021"
output: html_document
---

Simulating new dataset that is hierarchicaly structured

```{r, warning=FALSE, message=FALSE}
set.seed(456)
#Specifying population level values - true values that we are trying to estimate 

#Fixed effects
alpha_0 <-400 #intercept value 
beta_1 <-20 #slope 
sigma <- 5 #sd of response

#Specification of the between group variation 

# by-intercept sd, by_slope sd and correlation between intercept and slope
tau_0 <- 50 # by-group random intercet (countries)

tau_1 <- 10 # by-group random slope (countries)
rho <- -.3

n_countries<-10 #number of countries from which our babies are coming

countries_rfx <- faux::rnorm_multi(
  n=n_countries,
  mu = 0,
  sd = c(tau_0, tau_1),
  r = rho,
  varnames = c('T_0s','T_1s')
) #simulation of the multinormal distribution, where we have number of values equal to our countries. Mean is at 0 while we have two SD values for our intercept and slope adjustments 

babies_rfx=data.frame(T_0s=rep(countries_rfx$T_0s, each=20), T_1s=rep(countries_rfx$T_1s, each=20)) # We are going to repeat this information 20 times for each value, as we would like to sample 20 babies from each country

Babies <- data.frame(Country_id = rep(1:10, each=20),
babies_rfx) #Combining this in the new data frame, where we have id for each country as a first variable

#Simulation of other predictors
Babies$Age=round(runif(200,1,30))
Babies$Weight=rnorm(200,4000,500)
Babies$Gender=rbinom(200,1,0.5)
Babies$Gender=as.factor(Babies$Gender)
levels(Babies$Gender)=c('Girls','Boys')
#Simulation of the running distance where we specify all values on which distance depends. We have alpha_0 - intercept + T_0S - adjustments for intercept (country-level differences) + (beta_1 - slope + T_1s - adjustments for slope (country-level differences))*Babies$Age - our predictor, sigma - standard deviation for our values) 
Babies$RunningDist=rnorm(200,alpha_0 + Babies$T_0s+(beta_1+Babies$T_1s)*Babies$Age, sigma)
Babies$Country_id=as.factor(Babies$Country_id)
```

```{r}
table(Babies$Country_id) #How many observations we have for each group
```

Linear model: complete pooling

```{r}
mod1CP<-lm(RunningDist~Age, data=Babies)
summary(mod1CP)
```

Residual plot

```{r, fig.width=12, fig.height=5, fig.align='center'}
par(mfrow=c(1,1), bty='n',mar = c(5, 4, .1, .1), cex=1.1, pch=16)
plot(resid(mod1CP), ylab='Residuals', xlab='Index')
```

Linear model: no pooling

```{r}
mod1NP<-lm(RunningDist~Age+factor(Country_id)-1, data=Babies)
summary(mod1NP)
```

Residual plot

```{r, fig.width=12, fig.height=5, fig.align='center'}
par(mfrow=c(1,1), bty='n',mar = c(5, 4, .1, .1), cex=1.1, pch=16)
plot(resid(mod1NP), ylab='Residuals',xlab='Index')
```

Linear mixed-effect model: intercept adjustment

```{r, warning=FALSE, message=FALSE}
#install.packages('lme4')
require(lme4)

mult1<-lmer(RunningDist~Age+(1|Country_id), data=Babies) 
summary(mult1)
```

Linear mixed-effect model: slope for Age predictor adjustment and intercept adjustment for each country

```{r}
mult2<-lmer(RunningDist~Age+(1+Age|Country_id), data=Babies)
summary(mult2)
```

Fixed-effects:

```{r}
fixef(mult2)
```

Random effects 

```{r}
ranef(mult2)
```

No pooling model with Age*factor interaction:

```{r}
mod1NP2<-lm(RunningDist~Age*factor(Country_id)-1, data=Babies)
summary(mod1NP2)[4]
```

Residuals (no pooling versus mixed-effect model)

```{r, fig.width=12, fig.height=5, fig.align='center'}
par(mfrow=c(2,1), bty='n',mar = c(5, 4, .1, .1), cex=1.1,pch=16)
plot(resid(mult2))
plot(resid(mod1NP2))
```

Random intercepts and slopes adjustments for Age predictor for country factor

```{r}
mult2<-lmer(RunningDist~Age+(1+Age|Country_id), data=Babies)
print(summary(mult2), cor=F)
```

Statistical significance of the predictors: 

```{r, warning=FALSE, message=FALSE}
#install.packages('lmerTest')
require(lmerTest) #we need lmerTest and to rerun the model to get p-values
mult2<-lmer(RunningDist~Age+(1+Age|Country_id), data=Babies)
print(summary(mult2), cor=F)
```

Practical aspect

```{r, warning=FALSE, message=FALSE}
#install.packages('foreign')
require(foreign) #to read-in the data that are SPSS format (.sav) we need foreign package
mwell<-read.spss('data.sav', to.data.frame = T) #read.spss function and we specify that we should read it as a data frame
dim(mwell) #dimensions of our data (number of rows and columns)
mwell$Total=mwell$Watchwe_adj+mwell$Watchwk_adj+mwell$Comphwe_adj+mwell$Comphwk_adj+mwell$Smartwe_adj+mwell$Smartwk_adj #total amount of time spent watching screen in a week
```

Number of missing values in our outcome:

```{r}
table(is.na(mwell$mwb))
```

Numbe of missing values in our predictor:

```{r}
table(is.na(mwell$Watchwk))
```

Density plots for our variables

```{r, fig.width=12, fig.height=5, fig.align='center'}
par(mfrow=c(1,2), bty='n',mar = c(5, 4, .1, .1), cex=1.1,pch=16)
plot(density(mwell$mwb, na.rm=TRUE), main='')
plot(density(mwell$Total, na.rm = T), main='')
```

Subsetting the data - excluding NAs:

```{r}
mwell2=mwell[!is.na(mwell$mwb) & !is.na(mwell$Total),] # we are trying to subset our data frame and take only values that are not NAs. is.na is a boolean function that tells us whether one row is NA or not (TRUE or FALSE). !is.na indicates that we do not want TRUE na.values. We also have a logical parameter & that combines two conditions !is.na(outcome) & !is.na(predictor). Finally, we would like to filter our dataset by row and exclude all the rows that have either TRUE value in our predictor or our outcome. Therefore, we are looking at rows mwell[function goes here,] instead of mwell[,function goes here ] which would look at columns 
dim(mwell2) #dimensions of the smaller data
```

Scatter plot:

```{r, fig.width=12, fig.height=5, fig.align='center'}
cor(mwell2$mwb, mwell2$Total)
par(mfrow=c(1,1), bty='n',mar = c(5, 4, .1, .1), cex=1.1,pch=16)
plot(mwell2$Total[1:500], mwell2$mwb[1:500])
```

Number of observations for each of our potential random structures:

```{r}
table(mwell2$Ethnicg)
table(mwell2$REGION)
```

Building the model: 

```{r}
MWmod1<-lmer(mwb~(1|LANAME), data=mwell2) #random effect of Local area
MWmod2<-lmer(mwb~(1|LANAME)+(1|Ethnicg), data=mwell2) #crossed random effects of local area and ethnicity
MWmod3<-lmer(mwb~(1|REGION/LANAME)+(1|Ethnicg), data=mwell2) #nested random effect of local area that is nested in region and crossed with ethnicity
anova(MWmod1, MWmod2, MWmod3) #comparison of the models
```

Building the model: Fixed structure 1

```{r}
MWmod2a<-lmer(mwb~Total+(1|LANAME)+(1|Ethnicg), data=mwell2) #main effect of total
print(summary(MWmod2a), cor=F)
```

Building the model: Fixed structure 2

```{r}
MWmod2b<-lmer(mwb~Total+male+(1|LANAME)+(1|Ethnicg), data=mwell2) #main effect of total and main effect of sex
MWmod2c<-lmer(mwb~Total*male+(1|LANAME)+(1|Ethnicg), data=mwell2) #interaction between total and sex
anova(MWmod2a, MWmod2b, MWmod2c)#comparison of the models 
```

```{r}
print(summary(MWmod2c), cor=F)
```

Random structure: random slopes

```{r}
MWmod3c<-lmer(mwb~Total*male+(1+Total|LANAME)+(1|Ethnicg), data=mwell2) #random slopes for total predictor for Local area and intercept for local area 
```

```{r}
MWmod3c<-lmer(mwb~Total*male+(1|LANAME:male)+(1|Ethnicg), data=mwell2) #random intercept for unique combination between local area and sex 
anova(MWmod2c, MWmod3c) #comparison of the models
```

```{r}
MWmod3c<-lmer(mwb~Total*male+(1|LANAME)+(1|Ethnicg:male), data=mwell2) #random intercepts for unique combination between Ethnicity and sex
anova(MWmod2c, MWmod3c)
```

```{r}
summary(MWmod3)
```

Visualisation of the random structure:

```{r, warning=FALSE, message=FALSE, fig.width=12, fig.height=5, fig.align='center'}
require(sjPlot)
plot_model(MWmod3c, type='re', sort.est='sort.all', grid=FALSE)[1] #type='re' gives us random adjustments, while [1] indicates that we want only first plot
```

```{r , fig.width=12, fig.height=5, fig.align='center'}
plot_model(MWmod3c, type='re', sort.est='sort.all', grid=FALSE)[2]
```

Visualisation of the fixed effects (interaction): 

```{r, warning=FALSE, message=FALSE, fig.width=12, fig.height=5, fig.align='center'}
plot_model(MWmod3c, type='int')
```

Significance of the random effects

```{r}
ranova(MWmod3c)
```

Explained variance - R2

```{r, warning=FALSE, message=FALSE}
#install.packages('MuMIn')
require(MuMIn)
r.squaredGLMM(MWmod3c) # m stands for marginal, while c stands for conditional. Marginal is approximation of the explained variance by fixed-effect structure, while conditional is with both fixed and random-effect structure
```

Predictions of the model

```{r, fig.width=12, fig.height=5, fig.align='center'}
mwell2$predicted=predict(MWmod3c) #prediction values from model to our train dataset
par(mfrow=c(1,1), bty='n',mar = c(5, 4, .1, .1), cex=1.1,pch=16)
plot(mwell2$predicted, mwell2$mwb) #plot predicted versus observed data points
```
